\subsection{Overview and Context}
This thesis presents the design and implementation of a complete side scan sonar SLAM (SSS SLAM) system tailored for autonomous surface vessels (ASV's). The core structure of the work follows the flow of a full SLAM pipeline, beginning with a high level architecture and gradually moving toward detailed algorithmic and implementation specific components. The most important part of the thesis is the Side Scan Sonar SLAM Architecture chapter, which provides a unified overview of how all subsystems interact, from raw sonar acquisition to global trajectory estimation and map construction.
\\ \\
The chapters that follow expand each subsystem in depth. First, the Sonar Theory, Hardware, and System Modeling chapters establish the physical and mathematical foundations required for navigation, mapping, and estimation. Since SLAM performance depends directly on correct sensor models, motion models, and coordinate representations, this section defines the specific formulations used for the microAmpere ASV platform.
\\ \\
After establishing the required models and assumptions, the thesis transitions into detailed components of the SLAM system, state estimation, IMU preintegration, direct factors, local map generation, data association, optimizers, and global map construction. Each chapter describes the algorithms, mathematical tools, and design decisions used to form the complete SLAM pipeline.
\\ \\
Finally, a dedicated chapter outlines the frameworks and tools that will be used throughout the implementation. Since modern SLAM relies on well established libraries for factor graphs, optimization, vision processing, and middleware, this chapter summarizes the external software components that make the final system practical to build, deploy, and test on the microAmpere ASV.
