\subsection{MRPT Data Association and Perception Tools}
The Mobile Robot Programming Toolkit (MRPT) is a comprehensive open source C++ framework that provides essential tools for perception, mapping, localization, and robotic data processing. It is designed to support rapid development and testing of SLAM and navigation algorithms by offering a large collection of reusable software modules for sensing, motion modeling, and data fusion. MRPT is cross platform and integrates well with other robotics ecosystems such as ROS2, making it widely used in both academic and industrial research.  
\\ \\
In this work, MRPT is used primarily for data association and landmark management within the SLAM front-end. It includes advanced probabilistic data association algorithms such as the Joint Compatibility Branch and Bound (JCBB) method, which tests multiple measurement to landmark hypotheses simultaneously to ensure consistency across the entire observation set. This greatly reduces the risk of incorrect correspondences in cluttered or repetitive environments. MRPT also supports alternative association strategies, such as Nearest Neighbor (NN) and Mahalanobis distance based matching, allowing flexible adaptation to different sensor configurations and noise conditions.  
\\ \\
Beyond data association, MRPT offers a wide range of perception and robotics utilities. It includes libraries for point cloud processing, image handling, 3D visualization, occupancy grid mapping, particle filters, extended Kalman filters, and rawlog recording for dataset playback. These tools simplify tasks such as sensor calibration, dataset generation, and visualization of SLAM outputs. The framework also provides simulation environments, motion models for wheeled and differential robots, and extensive support for range and vision based sensors.  
\\ \\
Together, these features make MRPT a powerful framework for developing and validating SLAM front-ends. Its modular architecture allows selective integration of only the required components, reducing implementation time while maintaining flexibility and reliability. The official repository and related packages are available at \url{https://github.com/mrpt} \cite{MRPT_GitHub}.  
\\ \\
MRPT is distributed under the BSD-3-Clause license, permitting modification, redistribution, and research use while requiring acknowledgment of the original authors and inclusion of the license text in derived work.
