\subsection{OpenCV Computer Vision Library}
The Open Source Computer Vision Library (OpenCV) is a widely used open source framework for computer vision, image processing, and feature extraction. Originally developed by Intel in 1999 and now maintained by an active open source global community, it provides a comprehensive collection of algorithms for image analysis, geometric transformations, and visual perception tasks \cite{opencv_library}\cite{opencv_github}.  
\\ \\
OpenCV is implemented in C++ and designed for high performance visual computation, offering efficient data structures and hardware accelerated routines for real-time image processing. In this work, the C++ API is used primarily for the local map generation subsystem, where OpenCV supports both low-level image operations and high-level feature extraction. It is employed for filtering, thresholding, contour detection, and morphological transformations in probability maps, as well as for detecting and describing key features used in local mapping and spatial data association.  
\\ \\
The main motivation for using OpenCV lies in its mature, optimized implementation and extensive algorithmic coverage. Its core data types such as \texttt{cv::Mat}, \texttt{Point}, and \texttt{Rect} simplify the manipulation of image like probabilistic grids and map structures. These properties make it particularly suitable for handling local occupancy or likelihood maps, where visual operations directly translate to probabilistic updates. Moreover, its modular design and cross library interoperability allow seamless integration with other systems, such as ROS2 via \texttt{cv\_bridge}, and downstream state estimation frameworks like GTSAM and UKF-M for sensor fusion and visual SLAM.  
\\ \\
The official OpenCV website and source code repository are available at \url{https://opencv.org/} \cite{opencv_library} and \url{https://github.com/opencv} \cite{opencv_github}. The C++ library for OpenCV is distributed under the Apache-2.0 license, permitting unrestricted academic and commercial use, modification, and redistribution with appropriate attribution.  
\\ \\
In summary, OpenCV offers a robust, computationally efficient, and versatile foundation for image and map processing. Its optimized C++ implementation and comprehensive functionality make it an essential component for the perception and local mapping pipeline developed in this work.

