\subsection{Introduction}
Data association (DA) determines whether each new landmark observation corresponds to an existing map feature or should initialize a new one, making it one of the most critical components of SLAM. Its role is to ensure that the map remains internally consistent, every accepted match becomes a fixed constraint in the estimator, directly influencing all future pose and landmark updates \cite{DA_review}. Because a single incorrect association can corrupt the entire map or cause unbounded drift, SLAM relies on DA to filter out impossible or unlikely matches before they enter the optimization pipeline. This is achieved through strict statistical gating based on the Mahalanobis distance, which defines a confidence region in measurement space and limits the set of landmarks that must be considered for matching.
\\ \\
Most SLAM systems assume static or near static environments because reliably handling moving objects remains an open research challenge. Several approaches exist, robust estimators, non Gaussian noise models, motion removal, or explicit tracking, but none are universally reliable. For this thesis the seabed is treated as near static over short and medium time spans, which suits underwater mapping where sediment and structures change slowly. This simplifies DA to a geometric and probabilistic matching problem built on the extended range bearing model used throughout the thesis.
\\ \\
Every association that passes gating becomes a factor linking the current pose to a landmark, meaning the backend optimizer fully trusts the decision. To avoid catastrophic drift, the system relies solely on JCBB, using it in two modes depending on the situation. In ordinary local operation, tight gating leaves only 1 or very few candidates, causing JCBB to collapse naturally into a simple nearest neighbour outcome without performing full joint checks. In ambiguous cases or during loop closure attempts, JCBB switches to its full mode and evaluates joint compatibility using the full SLAM covariance, ensuring that only geometrically consistent multi landmark associations are accepted. Descriptor based loop closure triggering decides when this expensive global JCBB evaluation is necessary and when the lightweight local mode is sufficient.
\\ \\
Although many alternative DA frameworks exist, probabilistic methods like PDA/JIPDA, multi-hypothesis tracking, and mixture model formulations, they are too slow or fragile for real-time Side Scan Sonar SLAM (SSS SLAM) and are therefore only explored but not used in this thesis. The adopted pipeline instead couples gating, JCBB, and descriptor based loop closure detection into a unified scheme that in theory remains computationally efficient while still robust against false matches. This chapter outlines this DA strategy and motivates why these choices are best suited for reliable underwater SLAM.

