\subsection{Discussion}
\subsubsection{Real-Time Performance and Scalability}
The main result of this work is that the implemented SSS SLAM data-processing pipeline is able to run in real time on constrained embedded hardware. This is important because the goal of the system is not only to process side scan sonar data offline, but to provide a practical foundation for onboard use on autonomous underwater vehicles. The results show that the complete pipeline can process real sensor data, generate local maps, extract landmarks, and publish the required outputs while staying within the timing limits of the tested platform.
\\ \\
The tuned performance results are especially important. By reducing the map resolution and adjusting the feature extraction parameters, the local map generation pipeline becomes light enough to run with a much lower CPU and memory demand. This shows that the pipeline is not fixed to one specific hardware setup, but can be adapted to different compute budgets. On more powerful systems the parameters can be set to preserve more detail, while on smaller embedded platforms the same pipeline can be tuned to reduce runtime and memory usage. This makes the implementation scalable, since the same software structure can be reused across different vehicles and hardware platforms.
\\ \\
The results also show that the system is not only real-time capable in isolated algorithm tests, but also in a complete ROS2 setup where message passing, memory movement, publishing, and node scheduling are included. This is a more realistic result than only benchmarking individual functions, because onboard systems must share resources between perception, state estimation, SLAM, control, guidance, logging, and communication. The fact that the pipeline remains schedulable under these conditions means that the implementation is a practical step toward a deployable SSS SLAM system.

\subsubsection{Memory, Map Resolution, and Runtime}
The most important direct factor affecting runtime is the amount of map data that must be stored, processed, reconstructed, and published. In the local map generation results, changing the map resolution from $0.03\,\mathrm{m/pixel}$ to $0.06\,\mathrm{m/pixel}$ produces a large reduction in both runtime and memory usage. This confirms that map resolution is one of the most influential tuning parameters in the system. A coarser map contains fewer cells, which reduces the number of active chunks, the amount of persistent map state, the cost of dense map construction, the amount of gap filling, and the size of the published image.
\\ \\
This also explains why changing map resolution has a cascading effect through the whole pipeline. A smaller map is not only faster to generate, but also faster to filter, segment, label, and use for feature extraction. Therefore, reducing the map size improves several processing stages at once. This is why memory usage and runtime are tightly connected in the implemented system. When the dense map becomes large, the cost is not only computation, but also moving large amounts of memory and sending large ROS2 messages.
\\ \\
However, the resolution cannot be reduced without limit. If the map is made too coarse, useful spatial details may be lost, and the resulting landmarks may become less stable or less discriminative. The map resolution should therefore be understood as a practical tradeoff between retained sonar detail and available compute budget. The default configuration preserves more detail and already runs in real time, while the tuned configuration demonstrates how the system can be adapted when stricter embedded constraints are required.

\newpage

\subsubsection{State Estimation, Aiding Measurements, and Dead Reckoning}
A second important factor is the quality of the state estimate used during map generation. The map generation pipeline depends on accurate vehicle poses when inserting sonar swaths into the local map. When reliable aiding measurements are available, the state estimate remains stable, and the generated local maps stay spatially coherent. This improves both visual map quality and landmark repeatability.
\\ \\
When aiding measurements are lost, the system must rely more heavily on dead reckoning. During these periods, small pose errors accumulate over time. This causes sonar swaths to be inserted at slightly wrong positions, which can make the map spread, warp, or grow larger than necessary. The results show that this is not only a mapping quality problem, but also a performance problem. A degraded state estimate can increase the active map area, which increases memory usage and runtime.
\\ \\
This is where the full SLAM loop becomes important. Local maps are stable over short periods and can be used to extract repeatable landmarks. These landmarks can then be used for data association and loop closure, which can correct accumulated drift in the state estimate. A better state estimate then produces better local maps, which again produces more stable features. In this way, the system has the potential to form a positive feedback loop between mapping, feature extraction, data association, and state estimation.
\\ \\
The current results were produced without a fully closed loop SLAM backend correcting the trajectory from loop closures. Therefore, the full mission map still shows the expected effects of dead reckoning drift over long distances. This does not mean that the local map generator is unstable. Instead, it shows why local maps should be used as intermediate stable mosaics for feature extraction, and why loop closure is necessary before reconstructing the final global map. With reliable loop closures, the optimized trajectory can be used to place local maps into a global frame more consistently, giving a more stable full map than dead reckoning alone.

\subsubsection{Local Map Generation}
The local map generation algorithm works well on real side scan sonar data and is able to produce useful local reflectivity maps in real time. The implementation improves the practical use of probabilistic map generation by making the pipeline configurable, efficient, and suitable for embedded deployment. The sparse chunked map representation is especially important, because it avoids storing the whole dense mission scale map during processing. Instead, only active map regions are stored and updated, while old chunks can be removed/unloaded based on age.
\\ \\
The algorithm is still partly brute force in nature, since many candidate cells and beam interactions must be evaluated. For this reason, tuning is important. Parameters such as map resolution, pruning thresholds, chunk size, chunk age, and dense map update interval must be chosen according to the available compute budget and required map quality. Nevertheless, the results show that the current implementation is already more than efficient enough for real-time operation on Raspberry Pi 4 class hardware.
\\ \\
The most important outcome is therefore not only that the algorithm works, but that it has been converted into a practical and reusable software component. The implementation exposes the main tuning parameters, provides good default values, and is integrated into ROS2 while also keeping the core library separate from the ROS2 wrapper. This makes it easier to deploy the pipeline on any robot that already uses ROS2, while still allowing the raw processing libraries to be reused on custom systems that do not use ROS2.

\newpage

\subsubsection{Feature Extraction and Landmark Compression}
The purpose of feature extraction is not only to detect visible objects in the local map, but also to compress the map into a smaller and more useful representation for SLAM. A dense sonar map can become very large, especially over a full mission. Matching directly against the full map would be expensive, slow, and sensitive to local intensity changes. Instead, the system extracts landmarks and assigns each landmark a set of measurements, uncertainties, descriptors, and geometric properties.
\\ \\
This gives a much smaller representation of the environment. Instead of comparing large image maps during data association, the SLAM frontend can compare landmark descriptors and probabilistic weights. If two landmarks have similar descriptors and compatible geometry, they can be treated as possible association or loop closure candidates. This allows the system to prune the search space before more expensive geometric consistency checks are used.
\\ \\
In this sense, landmark extraction acts as a compression stage. The large local map is reduced to a compact feature database that contains the most relevant information for data association. This is important for scalability, because a global map image may grow to gigabytes, while the landmark database can remain much smaller and faster to search. Bounding boxes also provide an additional opportunity for compression. Nearby or overlapping detections could be merged into larger landmark objects when appropriate, reducing the number of individual features and improving matching speed.

\subsubsection{Feature Extraction Runtime and Height Estimation}
The feature extraction results show that the pipeline can run in real time, but also that some parts are much more expensive than others. Based on profiling, landmark height estimation is the largest single cost when enabled, using roughly $40\%$ of the feature extraction runtime. The initial filtering stage and the semantic segmentation stage each contribute roughly $20\%$, while the remaining operations together account for the final $20\%$. These values are approximate, but they clearly show where future optimization effort should be focused.
\\ \\
Height estimation is expensive because it requires additional shadow analysis for each detected landmark. The runtime difference is most visible during periods where many landmarks are detected. When there are few valid landmarks, height estimation has little work to do, and the fixed image processing stages dominate instead. This explains why disabling height estimation reduces runtime mainly in the parts of the dataset where many landmarks are present.
\\ \\
The current height estimation should also be understood as a rough approximation rather than a precise 3D reconstruction. It uses simplified sonar shadow geometry and depends on the quality of the detected shadow region. This can still provide a useful descriptor for landmark matching, but the uncertainty is high. Future work could improve this by using a more robust shadow model, better shadow segmentation, or a probabilistic height estimate. If height is not required by the later stages of navigation software stack, disabling it is a simple way to reduce runtime.

\newpage

\subsubsection{Data Association and Loop Closure}
The SSS data-processing pipeline is complete, but the full SLAM frontend was not fully completed in this work. Some initial experiments were performed with data association, loop closure detection, and descriptor based pruning. These experiments investigated whether landmark descriptors could be used to quickly reduce the number of possible association candidates before running a more expensive data association method. The preliminary results were promising, but the implementation was still experimental and was not mature enough to include as a final evaluated result.
\\ \\
This means that the most important next research step is data association and loop closure. The current work provides the required foundation, real-time local map generation, stable landmark extraction, descriptor assignment, uncertainty estimation, ROS2 integration, configurable parameters, and reusable libraries. Future work can therefore focus directly on the SLAM problem instead of rebuilding the full data-processing stack from scratch.
\\ \\
This is an important contribution of the work. A large amount of effort was spent on making the software clean, modular, documented, user friendly, configurable, and efficient. For ROS2 based vehicles, the pipeline can be deployed as ROS2 packages with only platform specific configuration. For non ROS systems, the core libraries can be used without the ROS2 wrappers. This makes the system easier to reuse on other underwater vehicles and gives future researchers a stronger starting point for implementing the full SLAM backend.

\subsubsection{Further Optimization}
Although the implemented system already runs in real time, there is still room for further optimization. One major bottleneck is memory movement. The results show that large dense maps are expensive not only because they require computation, but because they must be copied, reconstructed, transported, and published. Therefore, reducing memory overhead can improve runtime significantly.
\\ \\
One possible improvement is to use lossless compression for map data. Since lossless compression preserves the original information, it could reduce the amount of data moved through memory and communication layers without changing the map values. This would be especially useful for dense map images and intermediate map buffers, where the main cost is often data movement rather than mathematical computation. If processing can be performed directly on compressed or partially compressed representations, filtering and message passing could also become more efficient.
\\ \\
Another promising direction is GPU acceleration. The map generation problem has similarities to rasterization, illumination, and image reconstruction problems in classical computer graphics. Many operations are local and highly parallel, since different samples, pixels, or candidate cells can often be processed independently. This means that the algorithm could be well suited for shader based or general GPU implementations. Even small SBCs often include integrated GPUs, and using them could allow much larger maps or higher resolution sonar data to be processed without using several CPU cores.
\\ \\
The current CPU implementation is already real-time capable, which is an important baseline. A GPU implementation would not be required for the present dataset, but it could extend the method to higher fidelity sonars, larger missions, or stricter CPU budgets. This would allow the CPU to spend more time on data association, SLAM optimization, control, and guidance, while the GPU handles the dense map update and image style processing stages.

\newpage

\subsubsection{Summary}
Overall, the results show that the implemented SSS SLAM data-processing pipeline is a practical and reusable foundation for real-time sonar based SLAM. The pipeline can process real underwater data, generate local maps, extract landmarks, and run on constrained embedded hardware. The main remaining limitation is not the local data-processing pipeline, but the incomplete closed-loop SLAM frontend and backend. Reliable data association, loop closure, and graph optimization are required to turn the stable local maps and landmarks into a globally consistent SLAM solution.
\\ \\
The work therefore moves SSS SLAM from a mostly theoretical or offline concept toward a deployable embedded system. The local map generation and feature extraction pipeline is implemented, tested, documented, configurable, and ready to be used by future work. This allows later research to focus on the higher level SLAM problems, especially landmark association and loop closure, instead of spending time rebuilding the lower level sonar processing infrastructure.
